---
title: "analysis"
author: "Paul Melloy"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(RANN) # nearest neighbour
library(flextable)
library(officer)
library(data.table)
library(ggplot2)
library(mgcv) # gam models
library(here)
source(here("R/interpret_gam.R"))
```

```{r get_das}
calc_das <- function(d,days, Sd = "2022-03-31"){
  if(missing(days)){
    as.POSIXct(d, format = "%Y-%m-%d") - as.POSIXct(Sd, format = "%Y-%m-%d")
  }else{
      if(missing(d)){
        as.POSIXct(Sd, format = "%Y-%m-%d") + (days * 60 * 60 * 24)
      }
    }
}
calc_das("2022-05-15")
calc_das(days = 156)
```

### Read in data
```{r read_data}
cclo <- fread(here("output/2025-12-08_fulldf_cc_cercospoRa.csv"))[, "data" := "canopy_closure"]
eeeo <- fread(here("output/2025-12-08_fulldf_eo_cercospoRa_sus5.csv"))[, "data" := "epidemic_onset"]
```

We will combine the data so we can associate observation locations with canopy 
closure locations.  

### Inspect data
#### Check headings
```{r inspect_data_headings}
colnames(cclo)
colnames(eeeo)
```

We need to change the headings so they have the same names and join them.

#### Rename variables  
```{r rename_cols}
setnames(cclo,
         old = c("cc","area","treatment"),
         new = c("das","inoculum","fungicide"))

setnames(eeeo,
         old = c("area","treatment"),
         new = c("inoculum","fungicide"))

# Remove date column
eeeo[, date := NULL]
```

Join together the two datasets.  

```{r join}
dat <- rbind(cclo,eeeo)
```

#### Recheck variables    

```{r rename_vars}
unique(dat$fungicide)

unique(dat$inoculum)
# dat[inoculum == "roi", inoculum := "non-inoculated"]
# dat[inoculum == "ino", inoculum := "inoculated"]

unique(dat$source)
#dat[source == "uav", source := "uas"]


# Relevel fungicide factor
dat[,fungicide := factor(fungicide, levels = c("untreated", "treated"))]
dat[,inoculum := factor(inoculum, levels = c("non-inoculated", "inoculated"))]
```

<br>  

### Get nearest neighbour
Associate the closest observed epidemic onset date (EO) to the UAS or S2 
Earliest Estimated Epidemic Onset (EEEO) date.  

```{r nearest_neighbour}
obs_nearest <- nn2(data = dat[source == "observation",.(x,y)],
                   query = dat[,.(x,y)],
                   k = 2)

range(obs_nearest$nn.idx[,1])

# associate RS estimated EO with closest observation
dat[, obs_das:= dat[source == "observation",das][obs_nearest$nn.idx[,1]]]
```

<br>  
<br>  
***  

## Canopy Closure  

```{r subset_cclo}
cclo <- dat[data == "canopy_closure"]
```


Do a quick check on aov of the data to inspect average differences between 
inoculation and fungicide treatments.

```{r lm_check}
# check variance averages 
mod <- lm(das ~ source + inoculum + fungicide, data = cclo)
summary(mod)
```

### Plot data
```{r plot_data}
# plot EO data from each source an treatment
cclo |>
  ggplot(aes(source,das, fill = inoculum)) +
  geom_boxplot() +
  facet_wrap(~fungicide) +
  theme_bw() +
  labs(x = "Remote sensing platform",
       y = "Canopy Closure in days after sowing") +
  theme(legend.position = "top")+
  scale_fill_viridis_d(begin = 0.5)
```

### Model data  

#### Model 1  

> What are the average influences independent of other treatments accounting for
spatial variability?  

We will use Generalized Additive Models (GAM) to model the data, adding a spline
to account for the spatial variance across the field.  

```{r model_data1}
# what are the average influences independant of other treatments
mod1 <- gam(das ~ s(x,y) + source + inoculum + fungicide,
          data = cclo,
          method = "REML")
summary(mod1)
```

Check the spline fit to ensure no over fitting.  
```{r spline_fit1}
plot(mod1)
```

There seems an overfitting due to the experimental setup.
The spatial spline is accounting for too much of the inoculated areas.  

We will try again refitting with less knots.  

```{r model_data1a}
# what are the average influences independant of other treatments
mod1a <- gam(das ~ s(x,y, k=11) + source + inoculum + fungicide,
          data = cclo,
          method = "REML")
```

Check the spline fit to ensure no over fitting.  

```{r spline_fit1a}
plot(mod1a)
```

With fewer knots we have obtained a spatial correction for the field gradient.

Fungicide treated areas on average were not significantly different to untreated
for epidemic onset days.  

Non-Inoculated areas on average show earlier canopy closure, compared to 
inoculated areas.  

Remote sensing platform UAS detects average canopy closure 12 days earlier.  

<br>  

#### Model 2  

> Is there a significant interaction between measurement *source* and *inoculum*?  

```{r model_data2}
mod2 <- gam(das ~ s(x,y, k = 11) + source * inoculum + fungicide,
            data = cclo,
            method = "REML")
summary(mod2)
```

There is no significant interaction in inoculated areas and canopy closure dates
from different platforms.  

<br>  

#### Model 3  

> *Inoculum* and *fungicide* interaction?  

```{r model_data3}
mod3 <- gam(das ~ s(x,y, k = 11) + source + inoculum * fungicide,
            data = cclo)
summary(mod3)
```

There is a 
`r interpret_gam(mod3,coef = "inoculuminoculated:fungicidetreated" ,sig = TRUE)` interaction, 
between fungicide and inoculum 
(P = `r interpret_gam(mod3, 0,coef = "inoculuminoculated:fungicidetreated" ,sig = TRUE)`.  
Inoculated areas in fungicide treated plots show 
`r interpret_gam(mod3,5,coef = "inoculuminoculated:fungicidetreated",)` canopy closure date.

<br>  

#### Model 4  

> Is there a interaction between RS *source* and *fungicide*?  

```{r model_data4}
mod4 <- gam(das ~ s(x,y, k = 11) + source * fungicide + inoculum,
            data = cclo,
            method = "REML")
summary(mod4)
anova(mod4)
```


### Fungicide treatment  

Overall there was no significant effects of fungicide treated part of the field.
The only significant result showed remote sensing methods S2 and UAS resulted in
earlier EEEO dates in fungicide treated compared to untreated.  
The fungicide treated area seemed to show thicker canopy and I think this result
can be put down to spatial field effects.  

<br>  

#### Model 5  
Therefore the fungicide variable will be removed and this area treated as a 
field replicate.  

> Is there an interaction between RS platform and inoculated areas on canopy 
closure?  

```{r model_data5}
mod5 <- gam(das ~ s(x,y, k = 7) + source * inoculum,
            data = cclo[fungicide == "untreated"],
            method = "REML")
summary(mod5)
```

There is no canopy closure timing interaction between remote sensing platform 
and inoculated areas.  

##### results table  



```{r}
cclosure_estimates <- 
  data.table(Variable = c("Intercept","Super S2","UAS","Inoculated Area","Super S2 : Inoculated","UAS : Inoculated"),
             round(summary(mod5)$p.table,3))[,c(1,2,3,5)] |>
   flextable() |>
   align(align = "center", part = "all") |>
   fontsize(size = 8, part = "body") |>
   fontsize(size = 10, part = "header") |>
   italic(italic = TRUE, part = "header") |>
   set_caption(
      "Table ##: Estimated canopy closure in days after sowing (DAS) and the mean
      difference to the non-inoculated sentinal-2 data (intercept). 
      P values indicate statistical significance in comparison to the intercept."
   ) |>
   autofit()

cclosure_estimates
```


<br>  

#### Model 6  

```{r model_data6}
mod6 <- gam(das ~ s(x,y, k = 7) + source + inoculum,
            data = cclo[fungicide == "untreated"],
            method = "REML")
summary(mod6)
```


##### **Intercept**  
The intercept here represents: 
canopy closure dates in inoculated areas determined by sentinel-2 
and averages around `r interpret_gam(mod6, words = 0)` days
`r interpret_gam(mod6, words = 4)` sowing.  
  
##### **Remote sensing variables**  
UAS estimated epidemic onset an average of 
`r interpret_gam(mod6, words = 0, coeff = "sourceuas")` days 
`r interpret_gam(mod6, words = 5, coeff = "sourceuas")` than Sentinel 2.  

##### **Inoculation effect**

Non-inoculated areas were on average 
`r interpret_gam(mod6, words = 0, coeff = "inoculuminoculated")` days 
`r interpret_gam(mod6, words = 5, coeff = "inoculuminoculated")` 
than non-inoculated areas.  

As shown in `mod5` there is no interactive effect of source, indicating both 
remote sensing platforms showed inoculated areas had retarded phenology.

##### **Spatial effect**  
```{r}
plot(mod6)
```



Model comparisons

```{r}
anova(mod5,mod6)
```

No significant difference, use simpler model mod6.  


##### results table  



```{r}
icept <- round(unname(mod6$coefficients["(Intercept)"]),3)
tbl_dat <-  
  data.table(Variable = c("Intercept","Super S2","UAS","Inoculated Area"),
             round(summary(mod6)$p.table,3))[,c(1,2,3,5)][
               Estimate != icept, Estimate := Estimate + mod6$coefficients["(Intercept)"]]
cclosure_estimates6 <- 
  tbl_dat |>
  flextable() |>
   align(align = "center", part = "all") |>
   fontsize(size = 8, part = "body") |>
   fontsize(size = 10, part = "header") |>
   italic(italic = TRUE, part = "header") |>
   set_caption(
      "Table ##: Estimated canopy closure in days after sowing (DAS) and the mean
      difference to the non-inoculated sentinal-2 data (intercept). 
      P values indicate statistical significance in comparison to the intercept."
   ) |>
   autofit()

cclosure_estimates6
save_as_docx(cclosure_estimates6, path = "../output/cclosure_estimates.docx")
```


<br>  

<br>  
<br>  

***  

## Epidemic Onset  


```{r subset_eeeo}
eeeo <- dat[data == "epidemic_onset"]
```
<br>  

### Plot data
```{r plot_eo_data, fig.height= 8}
# plot EO data from each source an treatment
library(ggplot2)
eeeo[source != "observation",] |>
  ggplot(aes(x = obs_das, y = das, color = inoculum)) +
   geom_point(size = 4,alpha = .5,shape = 16) +
  facet_wrap(~fungicide + source) +
  theme_bw() +
  labs(title = "Epidemic onset by spatial location",
       x = "Observation",
       y = "UAS Estimated earliest EO") +
  theme(legend.position = "top")+
  xlim(c(90,180))+
  ylim(c(90,180))+
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dashed")+
  scale_color_viridis_d(begin = 0.4)
```

<br>  
<br>  

### Model data  

#### Model 1  

> What are the average influences independent of other treatments?  

We will use Generalized Additive Models (GAM) to model the data, adding a spline
to account for the spatial location of the data.  

```{r eo_model_data1}
# what are the average influences independant of other treatments
mod1_eo <- gam(das ~ s(x,y) + source + inoculum + fungicide,
          data = eeeo,
          method = "REML")
summary(mod1_eo)
```

Fungicide treated areas on average were not significantly different to untreated
for epidemic onset days.  

#### check spatial fit  

```{r eo_spatial_fit}
plot(mod1_eo)
```

This looks slightly over fit. 
We should use fewer knots.  

<br>  

#### Model 2  


```{r eo_mod2}
mod2_eo <- gam(das ~ s(x,y, k = 5) + source + inoculum + fungicide,
          data = eeeo,
          method = "REML")
plot(mod2_eo)
```

<br>  

#### Model 3  

> Is there a significant interaction between measurement source and inoculum?  

```{r eo_model_data3}
mod3_eo <- gam(das ~ s(x,y, k = 5) + source * inoculum + fungicide,
            data = eeeo,
            method = "REML")
summary(mod3_eo)
```

There is a 
`r interpret_gam(mod3_eo,1 ,"sources2:inoculuminoculated",sig = TRUE)` interaction 
between inoculated areas observed epidemic onset and remote sensing platform 
dates.  

<br>  

#### Model 4  


> Three-way interaction?  

```{r eo_model_data4}
mod4_eo <- gam(das ~ s(x,y, k = 5) + source * inoculum * fungicide,
            data = eeeo)
summary(mod4_eo)
```

As fungicide treated areas lead to 
`r interpret_gam(mod4_eo, 5 ,"fungicidetreated")` epidemic onset, 
however this was not observed with remote sensing platforms.

In the fungicide treated areas S2 showed 
`r interpret_gam(mod4_eo, 5 ,"sources2:fungicidetreated")` estimated epidemic onset, 
super S2 showed 
`r interpret_gam(mod4_eo, 5 ,"sources2s:fungicidetreated")` estimated epidemic onset, 
and USAS showed
`r interpret_gam(mod4_eo, 5 ,"sourceuas:fungicidetreated")` estimated epidemic onset, 
This was likely due to non-treated areas returning later canopy closure.  

Inoculated areas in fungicide treated plots showed for S2
`r interpret_gam(mod4_eo, 5 ,"sources2:inoculuminoculated:fungicidetreated")` estimated epidemic onset, 
super S2 showed 
`r interpret_gam(mod4_eo, 5 ,"sources2s:inoculuminoculated:fungicidetreated")` estimated epidemic onset, 
and USAS showed
`r interpret_gam(mod4_eo, 5 ,"sourceuas:inoculuminoculated:fungicidetreated")` estimated epidemic onset, 
compared to non-inoculated areas.



### Drop Fungicide Treatment  

The fungicide variable will be removed, initially I will test the model 
using fungicide as field replicate, then drop fungicide treated areas from the 
analysis.  

<br>  

#### Model 5  

```{r eo_model_data5}
mod5_eo <- gam(das ~ s(x,y, k =5) + source * inoculum,
            data = eeeo[fungicide != "treated"])
summary(mod5_eo)
```

##### **Intercept**  
The intercept here represents:  

> Observed EO dates in non-inoculated areas  

and averages around `r interpret_gam(mod5_eo, 0)` days 
`r interpret_gam(mod5_eo, 4)` sowing.  
  
##### **Remote sensing variables**  
Sentinel 2 estimated epidemic onset an average of 
`r interpret_gam(mod5_eo, 0,"sources2")` days 
`r interpret_gam(mod5_eo, 5,"sources2")` than observed EO.  

Super Resolution Sentinel 2 estimated earliest epidemic onset an average of 
`r interpret_gam(mod5_eo, 0,"sources2s")` days 
`r interpret_gam(mod5_eo, 5,"sources2s")` than observed EO.  

UAS estimated earliest estimated epidemic onset an average of 
`r interpret_gam(mod5_eo, 0,"sourceuas")` days 
`r interpret_gam(mod5_eo, 5,"sourceuas")` than observed EO.  

##### **Inoculation effect**  

Observed EO dates within inoculated areas were on average 
`r interpret_gam(mod5_eo, 0,"inoculuminoculated")` days 
`r interpret_gam(mod5_eo, 5,"inoculuminoculated")` 
days earlier than non-inoculated areas.  

However EEEO dates in inoculated areas produced by remote sensing methods 
S2 , super S2 and UAS were
`r interpret_gam(mod5_eo, 0,"sources2:inoculuminoculated")`,
`r interpret_gam(mod5_eo, 0,"sources2s:inoculuminoculated")` and 
`r interpret_gam(mod5_eo, 0,"sourceuas:inoculuminoculated")` days 
`r interpret_gam(mod5_eo, 5,"sourceuas:inoculuminoculated")` than the observed EO days.  

##### **Spatial Effect**  

```{r mod5_spatial}
plot(mod5_eo)
```

The spatial effect shows northern parts of the field showed earlier observed 
epidemic onset compared to Southern areas.


##### Results Table  

```{r}
var_names <- names(mod5_eo$coefficients)

icept <- round(unname(mod5_eo$coefficients["(Intercept)"]),3)

adj_coefs <- unname(
  c(mod5_eo$coefficients["(Intercept)"],
    mod5_eo$coefficients["(Intercept)"] + mod5_eo$coefficients[var_names[2]],
    mod5_eo$coefficients["(Intercept)"] + mod5_eo$coefficients[var_names[3]],
    mod5_eo$coefficients["(Intercept)"] + mod5_eo$coefficients[var_names[4]],
    mod5_eo$coefficients["(Intercept)"] + mod5_eo$coefficients[var_names[5]],
    mod5_eo$coefficients["(Intercept)"] + mod5_eo$coefficients[var_names[2]] + 
      mod5_eo$coefficients[var_names[5]] + mod5_eo$coefficients[var_names[6]],
    mod5_eo$coefficients["(Intercept)"] + mod5_eo$coefficients[var_names[3]] + 
      mod5_eo$coefficients[var_names[5]] + mod5_eo$coefficients[var_names[7]],
    mod5_eo$coefficients["(Intercept)"] + mod5_eo$coefficients[var_names[4]] + 
      mod5_eo$coefficients[var_names[5]] + mod5_eo$coefficients[var_names[7]]
  ))


tbl_dat <-  
  data.table(Variable = c("Intercept","Super S2","UAS","Inoculated Area"),
             round(summary(mod6)$p.table,3))[,c(1,2,3,5)][
               Estimate != icept, Estimate := Estimate + mod6$coefficients["(Intercept)"]]

eeeo_estimates <- 
  data.table(Variable = c("Intercept","Sentinal-2","Super Sentinal-2","UAS",
                          "Observed : Inoculated Area","S2 : Inoculated","Super S2 : Inoculated",
                          "UAS : Inoculated"),
             Estimate = round(adj_coefs,3),
             round(summary(mod5_eo)$p.table[,c(2,4)],3)) |>
   flextable() |>
   align(align = "center", part = "all") |>
   fontsize(size = 8, part = "body") |>
   fontsize(size = 10, part = "header") |>
   italic(italic = TRUE, part = "header") |>
   set_caption(
      "Table ##: Estimated epidemic onset in days after sowing (DAS) and the mean
      difference to the non-inoculated observed epidemic onset days (intercept). 
      P values indicate statistical significance in comparison to the intercept."
   ) |>
   autofit()

eeeo_estimates
save_as_docx(eeeo_estimates, path = "../output/earliest_onsets_estimates.docx")
```



<br>  

#### Model 6  

> Does the interaction still hold up when fungicide is excluded?  

```{r eo_model_data6}
mod6_eo <- gam(das ~ s(x,y, k = 5) + source + inoculum,
            data = eeeo[fungicide != "treated"])
summary(mod6_eo)
```


##### **Intercept**  
The intercept here represents: *Observed EO dates in non-inoculated areas*
and averages around `r interpret_gam(mod6_eo, 0)` days 
`r interpret_gam(mod6_eo, 4)` sowing.  
  
##### **Remote sensing variables**  
Sentinel 2 estimated epidemic onset an average of 
`r interpret_gam(mod6_eo, 0,"sources2")` days 
`r interpret_gam(mod6_eo, 5,"sources2")` than observed EO.  

Super Resolution Sentinel 2 estimated epidemic onset an average of 
`r interpret_gam(mod6_eo, 0,"sources2s")` days 
`r interpret_gam(mod6_eo, 5,"sources2s")` than observed EO.  

UAS estimated epidemic onset an average of 
`r interpret_gam(mod6_eo, 0,"sourceuas")` days 
`r interpret_gam(mod6_eo, 5,"sourceuas")` than observed EO.  

##### **Inoculation effect**

Observed EO dates within inoculated areas were on average 
`r interpret_gam(mod6_eo, 0,"inoculuminoculated")` days 
`r interpret_gam(mod6_eo, 5,"inoculuminoculated")` days 
than non-inoculated areas.  

##### **Spatial Effect**  
```{r}
plot(mod6_eo)
```

The spatial effect shows northern parts of the field showed earlier observed 
epidemic onset compared to Southern areas.


<br>  
<br>  

### Model comparisons

```{r model_comparisons}
anova(mod5_eo,mod6_eo)
AIC(mod5_eo,mod6_eo)
```

There is a significant difference between these models and model 5, with the
interaction shows a lower AIC.  

<br>  
<br>  

***  
## Proportional successes  

> What is the proportion of observed EO occuring later than EEEO?  

This should show us how successful the negative prognosis model is

```{r prop_success_dens}
tbl_prop_success <- 
    eeeo[source != "observation" & fungicide != "treated", 
      list("prop_success" = round(sum(das < obs_das)/NROW(das),3),
           n = .N), 
      by = .(source, inoculum, fungicide)][,.(source,
                                              inoculum,
                                              n,
                                              prop_success)] |>
  flextable() |>
  align_text_col(align = "center") |>
  set_header_labels(source = "RS method",
                    inoculum = "Inoculum",
                    prop_success = "proportional success") |>
   fontsize(size = 8, part = "body") |>
   fontsize(size = 10, part = "header") |>
   set_caption("Table ##: Proportion success of RS estimated onset dates occuring prior to spatially proximal observed epidemic onset")

tbl_prop_success
save_as_docx(tbl_prop_success, path = "../output/prop_success.docx")
```


Lets visualise the densities of the observed and estimated epidemic onset dates

```{r prop_success}
eeeo[, obs_dif := das - obs_das]

eeeo[source != "observation"] |>
  ggplot(aes(x = obs_dif, fill = inoculum)) +
  geom_density(alpha = .5) +
  facet_wrap(~source + fungicide) +
  theme_bw() +
  labs(x = "Difference between observed EO and Earliest estimated EO",
       y = "Density") +
  theme(legend.position = "top")+
  scale_fill_viridis_d()+
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 1)

```




